UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-06-03 ms/main.tex
more cleaning

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

%To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

@@

%\begin{figure}
%    \centering
%    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
%    \caption{A map of Great Britain coloured by a 3D UMAP projection. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour. Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres. A flattened 2D view of of the 3D projection used for colouring is presented in the top right. To protect participant privacy, data has been randomized as explained in the materials and methods.}
%    \label{fig:umap_ukbb_geo_map}
%\end{figure}

%\begin{figure}[h!]
%  \centering
%  \begin{subfigure}[b]{0.4\linewidth}
%    \includegraphics[width=\linewidth]{Figures/1KGP_UMAP_labels.jpeg}
%  \end{subfigure}
%  \rulesep
%  \begin{subfigure}[b]{0.4\linewidth}
%    \includegraphics[width=\linewidth]{Figures/1KGP_PC15_UMAP_labels.jpeg}
%  \end{subfigure}
%  \caption{UMAP embeddings of the 1KGP genotype data (left) and the top 15 principal components of the genotype data (right). Each point is coloured and labelled by its population. See Appendix~\ref{AppendixA} for label definitions.}
%  \label{fig:1KGP}
%\end{figure}

%\begin{figure}
%    \centering
%     \textbf{Test}\par\medskip
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=0.35\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
%     %\includegraphics[width=0.55\columnwidth]{images/meanWorldMap.jpg}
%    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
%    \label{fig:umap_ukbb_geo_map}
%    \end{subfigure}
%          \rulesep
%     \begin{subfigure}{\columnwidth}
%         \includegraphics[width=0.45\columnwidth]{images/meanWorldMap.jpg}
%   \caption{Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States. \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
%    \label{fig:umap_world_map}
%      \end{subfigure}
%      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \sgcomment{Merge captions. Make sure to have overall title! }}
%\end{figure}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{}
    \label{fig:umap_ukbb_geo_map}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/meanWorldMap.jpg}
    \caption{}
  \label{fig:umap_world_map}
  \end{subfigure}
  \caption{\textbf{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.} Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (a flattened projection is in the top right of \textbf{(a)}. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \textbf{(a)} Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres. \textbf{(b)} Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry. \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
\end{figure}

@@

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan\ref{fig:supp_fev_ridgeplots}.

These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).

@@

Identifying the best dimension reduction technique is to an extent subjective. However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}.  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

@@

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone.

%Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

Using UMAP on the UKBB data revealed a range of patterns of covariation between geography, phenotypes and genotypes, with traits such as height showing continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.

We found that pre-processing the data with PCA allowed for considerable time savings. Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}. In our experience, we find it useful to include as many PCs as are available.

@@

Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}. \sgcomment{Is this really a "material and method" comment? THe logical flow with the last comment is not clear to me.} \adpcomment{This was moved to trim some text from the main body, I can just remove it as it's redundant with info from 2 paragraphs earlier.}

@@

%%%%% boxplots end
\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
    \end{subfigure}
    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
    \label{fig:supp_umap_ukbb_worldmap_alt}
\end{figure*}

>>>>

\adpcomment{trim this pgh?}
Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{}
    \label{fig:umap_ukbb_geo_map}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/meanWorldMap.jpg}
    \caption{}
  \label{fig:umap_world_map}
  \end{subfigure}
  \caption{\textbf{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.} Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (a flattened projection is in the top right of \textbf{(a)}. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \textbf{(a)} Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres. \textbf{(b)} Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
\end{figure}

@@

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$). Distributions of FEV1 from Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.

%These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).

@@

Identifying the best dimension reduction technique is to an extent subjective. However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

@@

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone. With the UKBB data we revealed a range of patterns of covariation between geography, phenotypes and genotypes. Traits such as height showed continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showed sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.

We found that pre-processing the data with PCA allowed for considerable time savings. Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}. In our experience, we find it useful to include as many PCs as are available and computationally feasible.

@@

UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations.

@@

%%%%% boxplots end
%\begin{figure*}
%    \centering
%    \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
%    \end{subfigure}
%    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
%    \label{fig:supp_umap_ukbb_worldmap_alt}
%\end{figure*}
